A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-07 tex/sec3-defs.tex
[ tex ] Explain mode-correctness

% We may confine the use of an arbitrary variable set $\Xi$ to some set in the family of sets $\Fin_n$ of natural numbers less than $n$, as defined by \citet{Dybjer1994}, for the following reasons.
% In bidirectional type synthesis, a decidable equality on $\Type_{\Sigma}(\Xi)$ is necessary, as we need to compare the synthesised type $B$ with a given type $A$ for the $\Rule{Sub}$ rule.
% If $\Xi$ has a decidable equality as $\Fin_n$ does, then $\Type_{\Sigma}(\Xi)$ also has a decidable equality.
% In addition, the arity of a construct is finite, so it suffices to use natural numbers to denote type variables in a type.
% As we will only consider a finite collection of types (in a typing rule or in a signature), we can formally restrict $\Xi$ to the family of $\Fin_n$ in later definitions.
% For illustrative purposes, though, we still use named variables for presentation.

@@

A simply typed language consists of
\begin{enumerate*}
  \item a set of raw terms $\isTerm{t}$ indexed by a set $V$ of untyped variables where each construct is allowed to bind some variables like $\Rule{Abs}$ and to take multiple arguments like $\Rule{App}$;
  \item a set of typing derivations indexed by a list $\Gamma$ of typed variables, a type~$A$, and a raw term $\isTerm{t}$ to set their type constraints. 
\end{enumerate*}
Therefore, to specify a term construct, we enrich the notion of arity with some sort and extension context for variable binding:
\begin{definition}\label{def:binding-arity}
  A \emph{binding arity} with a sort $T$ is an inhabitant of $\left(T^* \times T\right)^* \times T$ where $T^*$ is the set of lists over $T$.
  In a binding arity ${[\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n}} \to {A_0}$, every $\Delta_i$ and $A_i$ refers to the \emph{extension context} and the sort of the $i$-th argument, respectively; for brevity, we denote it as ${[\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n}} \to {A_0}$ and omit $[\Delta_i]$ if $\Delta_i$ is empty.
\end{definition}

@@

Next, akin to a signature, our binding signature $\Omega$ consists of a set of operation symbols along with their respective binding arities.
It also includes additionally the set $\Xi$ of type variables that appear in the binding arity with the sort $\Type_{\Sigma}(\Xi)$ for each operation:
\begin{definition}\label{def:binding-signature}
  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
  \[
    \arity(o) \colon O \to \sm{\Xi : \UU} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
  \]
\end{definition}
That is, each inhabitant $o: O$  is associated with a set $\Xi$ of type variables and a binding arity with sort $\Type_{\Sigma}(\Xi)$.
We may simply call $\arity(o)$ the arity of $o$, denoted by $\aritysymbol{o}{\Xi}{[\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n}}{A_0}$.
By a \emph{simply typed language} $(\Sigma, \Omega)$, we mean a pair of a type signature $\Sigma$ and a binding signature $\Omega$.

>>>>

@@

A simply typed language consists of
\begin{enumerate*}
  \item a set of raw terms $\isTerm{t}$ indexed by a set $V$ of untyped variables where each construct is allowed to bind some variables like $\Rule{Abs}$ and to take multiple arguments like $\Rule{App}$;
  \item a set of typing derivations indexed by a list $\Gamma$ of typed variables, a type~$A$, and a raw term $\isTerm{t}$ to set their type constraints. 
\end{enumerate*}
Therefore, to specify a term construct, we enrich the notion of arity with some sort and extension context for variable binding:
\begin{definition}\label{def:binding-arity}
  A \emph{binding arity} with a sort $T$ is an inhabitant of $\left(T^* \times T\right)^* \times T$ where $T^*$ is the set of lists over $T$.
  In a binding arity ${[\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n}} \to {A_0}$, every $\Delta_i$ and $A_i$ refers to the \emph{extension context} and the sort of the $i$-th argument, respectively; for brevity, we denote it as ${[\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n}} \to {A_0}$ and omit $[\Delta_i]$ if the extension context $\Delta_i$ is empty.
\end{definition}

@@

Next, akin to a signature, our binding signature $\Omega$ consists of a set of operation symbols along with their respective binding arities.
It also includes additionally the set $\Xi$ of type variables that may appear in the binding arity with the sort $\Type_{\Sigma}(\Xi)$ for each operation:
\begin{definition}\label{def:binding-signature}
  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
  \[
    \arity(o) \colon O \to \sm{\Xi : \UU} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
  \]
\end{definition}
That is, each inhabitant $o: O$  is associated with a set $\Xi$ of type variables and a binding arity with sort $\Type_{\Sigma}(\Xi)$.
We may simply call $\arity(o)$ the arity of $o$, denoted by $\aritysymbol{o}{\Xi}{[\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n}}{A_0}$.
By a \emph{simply typed language} $(\Sigma, \Omega)$, we mean a pair of a type signature $\Sigma$ and a binding signature $\Omega$.

### 2023-07-07 tex/sec4-pre-synthesis.tex
[ tex ] Explain mode-correctness

\begin{theorem}\label{lem:soundness-completeness}
$\Gamma |-_{\Sigma, \Omega} t :^\dir{d} A$ if and only if\/ $\erase\Gamma |-_{\Sigma, \Omega} t^\dir{d}$ and\/ $\Gamma |-_{\Sigma, \erase\Omega} t : A$.
\end{theorem}

>>>>

\begin{theorem}\label{lem:soundness-completeness}
$\Gamma |-_{\Sigma, \Omega} t :^\dir{d} A$ if and only if both\/ $\erase\Gamma |-_{\Sigma, \Omega} t^\dir{d}$ and\/ $\Gamma |-_{\Sigma, \erase\Omega} t : A$.
\end{theorem}

### 2023-07-07 tex/sec5-synthesis.tex
[ tex ] Explain mode-correctness

\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}
This section is devoted to defining mode-correctness and deriving bidirectional type synthesis for any mode-correct bidirectional type system $(\Sigma, \Omega)$.
\Cref{sec:mode-correctness} formalises the concept of mode-correctness and establishes the uniqueness of synthesised types.
This indicates that for a given raw term and context, any two synthesised types must be equal.
This characteristic is also used in the proof of the logical decidability of bidirectional type synthesis.
Later, in \Cref{subsec:trichotomy}, we introduce the decidability of bidirectional synthesis but postpone its proof until \Cref{subsec:bidirectional-synthesis-checking}.
At that point, we will combine mode preprocessing, soundness, and completeness from \Cref{sec:pre-synthesis} to show the trichotomy on raw terms as a simple corollary.
Note that the result is applicable to an arbitrary mode-correct bidirectional type system, rather than being restricted to the simply typed $\lambda$-calculus as we did in \Cref{sec:key-ideas}.

\subsection{Mode Correctness}\label{sec:mode-correctness}
As \citet[Section~3.1]{Dunfield2021} outlined, mode-correctness for a bidirectional type system means that for every bidirectional typing rule:
\begin{enumerate*}
\item each `input' type variable in a premise must be synthesised from `earlier' premises or, if the mode of the conclusion is checking, provided by the conclusion;
\item each `output' type variable in the typing judgement of the conclusion should have been synthesised from premises already, if the mode of the conclusion is synthesis.
\end{enumerate*}
An `input' type variable refers to types in an extension context and the type of a premise in the checking mode.
It is important to note that the order of premises in a bidirectional typing rule matters, as it also sets the order of bidirectional type synthesis.

Consider the rule $\ChkRule{Abs}$ (\Cref{fig:STLC-bidirectional-typing-derivations}) as an example.
This rule is mode-correct, as the type variables $A$ and $B$ in its only premise are already provided by its conclusion $A \bto B$.
Likewise, the rule $\SynRule{App}$ for an application term $\isTerm{t\;u}$ is mode-correct because:
\begin{enumerate*}
\item the type $A \bto B$ of the first argument $t$ is synthesised, thereby ensuring type variables $A$ and $B$ must be known if successfully synthesised;
\item the type of the second argument $u$ is checked against $A$, which was synthesised earlier;
\item as a result, the type of an application $t\;u$ can be synthesised.
\end{enumerate*}

As this is a condition for \emph{typing rules} and not for raw terms or typing derivations, this property \emph{cannot} be formalised without the concept of bidirectional binding signature presented in \Cref{sec:defs}.
This contrasts with the properties in \Cref{sec:pre-synthesis}, which are properties that can still be expressed for individual systems even without the general definition of bidirectional type systems.
Now that we have the notion of signature available, we can articulate mode-correctness more precisely.

First, let us introduce the finite subset\footnote{%
  There are various definitions for finite subsets of a set within Martin-L\"{o}f type theory.
  However, for our purposes, the choice among these definitions is not a matter of concern.
}
$\fv(A)$ and $\fv(\Gamma)$ of (free) variables in a type and in a context, respectively, which are defined in the usual way:
\begin{enumerate*}
\item For a type $A : \Type_{\Sigma}(\Xi)$ define the subset $\fv(A)$ of $\Xi$ inductively by $\fv(x) = {x}$ and $\fv(\tyOp_{i}(A_1, \ldots, A_n)) = \bigcup_{i = 1}^{n} \fv(A_i)$;
\item by reusing this notation, for a context $\Gamma : \Type_{\Sigma}(\Xi)$ define the set $\fv(\Gamma)$ as $\fv(\emptyset) = \emptyset$ and $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
\end{enumerate*}
As we have sketched informally, the mode-correctness condition for a synthesis rule is different from a checking rule
\begin{definition}[Mode-Correctness] \label{def:mode-correctness}
  An operation $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}}}{A_0^{\dir{d}}}$ is \emph{mode-correct} whenever
  \begin{enumerate}
    \item if $d$ is $\syn$ (synthesis);
    \item if $d$ is $\chk$ (checking);
  \end{enumerate}

\end{definition}

>>>>

\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}
This section is devoted to defining mode-correctness and deriving bidirectional type synthesis for any mode-correct bidirectional type system $(\Sigma, \Omega)$.
\Cref{sec:mode-correctness} formalises the concept of mode-correctness and establishes the uniqueness of synthesised types.
This indicates that for a given raw term and context, any two synthesised types must be equal.
This property will also be used in the proof of the logical decidability of bidirectional type synthesis.
Later, in \Cref{subsec:trichotomy}, we introduce the decidability of bidirectional synthesis but postpone its proof until \Cref{subsec:bidirectional-synthesis-checking}.
At that point, we will combine mode preprocessing, soundness, and completeness from \Cref{sec:pre-synthesis} to show the trichotomy on raw terms as a simple corollary.

\subsection{Mode Correctness}\label{sec:mode-correctness}
As \citet{Dunfield2021} outlined, mode-correctness for a bidirectional type system means that for every typing rule:
\begin{enumerate*}
\item each `input' type variable in a premise must be synthesised from `earlier' premises or provided by the conclusion, if the rule is checking;
\item each `output' type variable in the conclusion should have been synthesised from premises already, if the rule is synthesis.
\end{enumerate*}
Here an `input' type variable refers to types in an extension context and the type of a premise in the checking mode.
It is important to note that the order of premises in a bidirectional typing rule matters, as it also sets the order of bidirectional type synthesis.

Consider the rule $\ChkRule{Abs}$ (\Cref{fig:STLC-bidirectional-typing-derivations}) as an example.
This rule is mode-correct, as the type variables $A$ and $B$ in its only premise are already provided by its conclusion $A \bto B$.
Likewise, the rule $\SynRule{App}$ for an application term $\isTerm{t\;u}$ is mode-correct because:
\begin{enumerate*}
\item the type $A \bto B$ of the first argument $t$ is synthesised, thereby ensuring type variables $A$ and $B$ must be known if successfully synthesised;
\item the type of the second argument $u$ is checked against $A$, which has been synthesised earlier;
\item as a result, the type of an application $t\;u$ can be synthesised.
\end{enumerate*}

As this is a condition for \emph{typing rules} and not for raw terms or typing derivations, this property \emph{cannot} be formalised without the concept of bidirectional binding signature presented in \Cref{sec:defs}.
This contrasts with the properties in \Cref{sec:pre-synthesis}, which are properties that can still be expressed for individual systems even without a general definition of bidirectional type systems.
Now that we have the notion of signature available, we can articulate mode-correctness more precisely.

But, first, we assume that the set $\Xi$ of variables has a \emph{decidable equality}, rather than being an arbitrary set, so that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.
This requirement for decidable equality is necessary, as type comparisons are ubiquitous throughout type synthesis.
To simplify our choice, we could just restrict $\Xi$ to any set within the family of sets $\Fin_n$ of natural numbers less than $n$, as defined by \citet{Dybjer1994}.
Given that this set has a decidable equality and the arity of a type construct is always finite, it would be a viable choice.
For clarity in presentation, however, we will continue to use named variables.

Now we proceed with the definition of mode-correctness.
As we have outlined at the beginning, the condition of mode-correctness for a synthesis rule is different from that of a checking rule, and the argument order also matters.
Defining the condition directly for a signature can be somewhat intricate.
Hence, we instead choose to define the condition in a few more steps, based on the list of arguments---more specifically, triples of an extension context $\Delta_i$, a type $A_i$, and a mode $\dir{d_i}$---pertaining to an operation, and subsequently on a signature.
We also need some auxiliary definitions for the subset of variables of a type and the set of variables that have been synthesised:
\begin{definition}
  The finite subset\footnote{%
  There are various definitions for finite subsets of a set within Martin-L\"{o}f type theory.
  However, for our purposes, the choice among these definitions is not a matter of concern.}
  of \emph{variables} which occur in a type $A$ is denoted by $\fv(A)$.
  The finite subset $\fv(\Gamma)$ of variables in a context $\Gamma$ is defined inductively by $\fv(\cdot) = \emptyset$ and $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
  For a list of \emph{arguments} $\overrightarrow{[\Delta_i]A_i^{d_i}}$, define the set of \emph{synthesis type variables} inductively as 
  \begin{align*}
    \synvar(\cdot)                                  & = \emptyset  \\
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\chk}}) & = 
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \\
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\syn}}) & = 
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \cup \fv(A_{n}).
  \end{align*}
\end{definition}
This subset accumulates type variables of a synthesising argument and they are exactly those type variables that would be synthesised during type synthesis.

\begin{definition}\label{def:mode-correctness}
  The \emph{mode-correctness for an argument list} $\Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}}$  with respect to a subset $S$ of $\Xi$ is defined inductively as
  \begin{align*}
    \MC_{\mathit{as}}( \cdot ) ={} & \top \\
    \MC_{\mathit{as}}(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\syn}) 
    & = \MC_{\mathit{as}}(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})
    \land \fv(\Delta) \subseteq \left( S \cup \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \right)\\
    \MC_{\mathit{as}}(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\chk})
    & = \MC_{\mathit{as}}(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})
    \land \fv(\Delta, A_n) \subseteq \left( S \cup \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})\right).
  \end{align*}
\end{definition}
This definition encapsulates the idea that every `input' type variable, possibly deriving from a context $\Delta$ or a checking argument $A_n$, must be an 'output' variable from $\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ or, if the rule is checking, belong to the set $S$ of 'input' variables in its conclusion.
This condition must be met for every tail of the argument list to ensure that `output' variables accessible at each argument position are from preceding arguments only.
\begin{definition}
  An operation $o\colon \Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}} \to A_0^{\dir{d}}$ is \emph{mode-correct} provided that
  \begin{enumerate}
    \item $d$ is $\chk$, its argument list is mode-correct w.r.t.\ $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ contains every inhabitant of $\Xi$;
    \item $d$ is $\syn$, its argument list is mode-correct w.r.t.\ $\emptyset$, and the finite subset $\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
  \end{enumerate}
  A bidirectional binding signature $\Omega$ is \emph{mode-correct} if its operations are all mode-correct.
\end{definition}
Finally, for a checking operation, an 'input' variable of an argument could be derived from $A_0$, as these are known during type checking as an input.
Since every inhabitant of $\Xi$ can be located in either $A_0$ or synthesis variables, we can determine a concrete type for each instance of $\Xi$ during type synthesis.
On the other hand, for a synthesis operation, we do not have any known variables at the onset of type synthesis, so the argument list should be mode-correct with respect to $\emptyset$.
Also, the set of synthesis variables alone should include every type variable in $\Xi$ and particularly in $A_n$.
